Implementation and assessment of tools for integrating forest roads into projections of the cumulative effects of disturbance on wildlife

### 2023-04-26 analysis/paper/paper.Rmd
writing in progress

While road projection tools are not typically used in forest landscape models, they are important for harvest planning, and methods [@anderson_projecting_2004; @clark_three-stage_2000; @dean_finding_1997; @epstein_combinatorial_2006; @meignan_heuristic_2012; @stuckelberger_improved_2007] and tools [e.g. @remsoft_remsoft_2019; @spatial_planning_systems_patchworks_nodate] exist to accomplish this task in an operational context. These have occasionally been incorporated into forest landscape models [@leston_quantifying_2020], however they were not designed for this purpose. The algorithms contained in proprietary closed-source software are difficult to assess or fully integrate into models that include stochastic disturbances [@sturtevant_understanding_2021; @de_pellegrin_llorente_recognizing_2017] and the level of detail required to parameterize them is often difficult to obtain at landscape scales. Harvest road planning tools likely produce the most accurate road projections but the cost of preparing and running them can be prohibitive in contexts where a rougher approximation of road locations may suffice.

@@

```{r study-area, fig.cap="Revelstoke timber supply area in British Colombia, Canada. The colour shows the cost of building roads across each pixel.", cache=TRUE}
tsaBoundary <- read_sf(here(data_path_raw, "tsa27_boundaries.gpkg"))
#cost surface raster layer
tsaCost <- terra::rast(here(data_path_drvd, "input_cost.tif"))

@@

Hardy's method [-@hardy_forest_2019;-@hardy_landis-ii_2023] is similar to our DLCP method in that it breaks the MTAP down into many STAPs and finds the least cost path for each with the ordering of road construction determined by the distance from the existing road network. One difference is that Hardy uses the Dijkstra algorithm to solve a “single-source many-targets shortest path problem”, which is a special case of STAPs that finds the path between the harvested cell and one of multiple potential connection points on the road network. On the other hand, our algorithm first finds the closest point on the road network to the selected landing point and then uses the Dijkstra algorithm to solve least cost path between the two. However, since the cost of construction on existing roads is zero we expect that in practice these two approaches are similar and the algorithm will select the least cost location for the road to connect to the existing road network. It is possible that a sub-optimal road connection location is chosen by our algorithm when the existing road network is not fully connected and the closest point is on an existing road fragment that is not connected to the optimal part of the road network. Hardy's method also has a different approach for building roads inside cutblocks, it iterates over each cell in the harvest area and builds a road to it if it is not within the skidding distance of an existing road. In our algorithm we create landing points in the harvested area to serve as the starting points for the roads to be built based on an expected landing density.

<!--
TO DO: should this paragraph about speed and memory go in the results or discussion rather than methods?
-->
The MST method is expected to be closer to an optimal solution than either of the LCP methods described above but at the cost of being more computationally intensive and taking longer to solve. The Hardy QGIS plugin was much faster to run than our R package and more robust to analyzing large landscapes. The `igraph` package that we used to implement the Dijkstra algorithm requires the values of the cost raster to be read into memory in R which can lead to high memory usage on larger landscapes. See Appendix X for a more detailed exploration of the speed and memory requirements of the different methods.

@@

```{r proj-meth-fig, fig.cap="We compared minimum spanning tree (MST), dynamic least cost paths (DLCP) and the Hardy QGIS plugin road network projection algorithms. Landing locations were selected as the centroid of the cutblock, or a random or regularly spaced sample of points inside each cutblock (Table 2). Target cutblocks are outlined in grey, existing roads are black, projected roads are yellow, observed roads are red. Cost (1 ha resolution) is indicated by grey shading, with darker grey indicating higher cost. The large four sided cutblock in the top right was added to give a better example of the algorithms behaviour inside cutblocks."}
knitr::include_graphics(here("analysis/figures/projection_methods_figure.png"))
```

@@

```{r aspat-perf-fig, fig.height=4, fig.width=fig_widths["two"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection method variants with a 1 ha resolution cost raster. Metrics were summarised within cutblocks, outside cutblocks, and overall. Values greater than 0 indicate overprojection of the metric and negative values indicate an underprojection. Cutblocks that did not intersect the observed road network were excluded"}
# organize data for ggplot
doAspatPlot <- function(tab){
  meanTable_long <- tab %>% dplyr::select(-runTime, -order) %>% 
  pivot_longer(-c(sampleType, sampleDens, areaMean, method),
               names_to = "metric", values_to = "response")%>%
  mutate(sampleType = ifelse(sampleType == sampleDens, NA_character_, 
                             sampleType),
         sampleDens = paste(method, sampleType, sampleDens) %>% 
           factor(levels = c("mst NA centroid",
                             "NA NA observed",
                             "mst random low sample density",
                             "mst regular low sample density",
                             "mst random high sample density",
                             "mst regular high sample density",
                             "dlcp regular high sample density",
                             "NA NA klementQGIS",
                             "NA NA NA"),
                  labels = c("MST centroid", 
                             "Observed",
                             "MST random low density",
                             "MST regular low density",
                             "MST random high density",
                             "MST regular high density",
                             "DLCP regular high density",
                             "Hardy QGIS",
                             "Cutblocks only")),
         metric = factor(metric,
                         levels = c("roadDensityMean", "roadPresenceMean",
                                    "distanceToRoadMean",
                                    "forestryDisturbanceMean",
                                    "roadDisturbanceMean"),
                         labels = c("Road\ndensity", "Road\npresence",
                                    "Distance\nto road",
                                    "Forestry\ndisturbance\nfootprint",
                                    "Road\ndisturbance\nfootprint")))

@@

```{r proj-map-fig, fig.height=9, fig.width=fig_widths["two"], fig.cap="Comparison of projection results for the minimum spanning tree projection with regular high density sampling to the observed road network. The primary map shows the observed distance to the nearest road minus the projected distance and the insets show close-ups of the road networks with satellite imagery to give more context. a) Projected and observed road networks follow similar patterns although projected roads cross valley bottoms more frequently. b) An example of a misidentified cutblock. c) Projected roads do not match observed switchback roads.  The satellite images are from Google (2022)."}

@@

```{r spat-perf-fig, fig.height=4.5, fig.width=fig_widths["two"], fig.cap="Comparison of performance metrics for each projection method variant using the 1 ha resolution cost raster. F-measure is the harmonic mean of precision and sensitivity, precision is the proportion of predicted roads that are observed roads and sensitivity is the proportion of observed roads that were correctly predicted. Values closer to one indicate better performance. Cutblocks that did not intersect the observed road network were excluded"}
doSpatPerf <- function(tab){
  matchData_sum <- tab %>% ungroup() %>%
    mutate(agreement = factor(agreement,
                              levels = c("False negative", "False positive",
                                         "Agree roaded","Agree roadless",
                                         "Pre-existing roads")),
           sampleDens = paste(method, sampleType, sampleDens) %>%
             str_replace("1e-06", "low density") %>%
             str_replace("1e-05", "high density") %>%
             str_replace("centroid low density", "centroid") %>%
             str_replace("klementQGIS NA", "klementQGIS") %>%
             factor(levels = c("mst centroid",
                               "mst random low density",
                               "mst regular low density",
                               "mst random high density",
                               "mst regular high density",
                               "dlcp regular high density",
                               "NA klementQGIS", 
                               "NA cutOnly NA"),
                    labels = c("MST centroid", 
                               "MST random low density",
                               "MST regular low density",
                               "MST random high density",
                               "MST regular high density",
                               "DLCP regular high density",
                               "Hardy QGIS", 
                               "Cutblocks only")),
           metric = factor(metric,
                           levels = c("roadPresence",
                                      "forestryDisturbance",
                                      "roadDisturbance"),
                           labels = c("Road presence",
                                      "Forestry disturbance\nfootprint",
                                      "Road disturbance\nfootprint"))) %>%
    group_by(metric, sampleDens, agreement) %>%
    summarise(count = sum(count), .groups = "drop_last") %>%
    mutate(perc = count/sum(count)*100)

>>>>

While road projection tools are not typically used in forest landscape models, they are important for harvest planning, and methods [@anderson_projecting_2004; @clark_three-stage_2000; @dean_finding_1997; @epstein_combinatorial_2006; @meignan_heuristic_2012; @stuckelberger_improved_2007] and tools [e.g. @remsoft_remsoft_2019; @spatial_planning_systems_patchworks_nodate] exist to accomplish this task in an operational context. These have occasionally been incorporated into forest landscape models [@leston_quantifying_2020] but they were not designed for that purpose. The algorithms contained in proprietary closed-source software are difficult to assess or fully integrate into models that include stochastic disturbances [@sturtevant_understanding_2021; @de_pellegrin_llorente_recognizing_2017] and the level of detail required to parameterize them is often difficult to obtain at landscape scales. Harvest road planning tools may produce more accurate road projections but the cost of preparing and running them can be prohibitive in contexts where a rougher approximation of road locations may suffice.

@@

```{r study-area, fig.cap="Road construction costs according to a simple model informed by slope and major barriers (Table \@ref(tab:cost-tbl)) in the Revelstoke timber supply area in British Colombia, Canada.", cache=TRUE}
tsaBoundary <- read_sf(here(data_path_raw, "tsa27_boundaries.gpkg"))
#cost surface raster layer
tsaCost <- terra::rast(here(data_path_drvd, "input_cost.tif"))

@@

Hardy's method [-@hardy_forest_2019;-@hardy_landis-ii_2023] is similar to our DLCP method in that it breaks the MTAP down into many STAPs and finds the least cost path for each with the ordering of road construction determined by the distance from the existing road network. One difference is that Hardy uses the Dijkstra algorithm to solve a “single-source many-targets shortest path problem”, which is a special case of STAPs that finds the path between the harvested cell and one of multiple potential connection points on the road network. In contrast our algorithm first finds the closest point on the road network to the selected landing point and then uses the Dijkstra algorithm to solve least cost path between the two. However, since the cost of construction on existing roads is zero we expect that in practice these two approaches are similar and the algorithm will select the least cost location for the road to connect to the existing road network. It is possible that a sub-optimal road connection location is chosen by our algorithm when the existing road network is not fully connected and the closest point is on an existing road fragment that is not connected to the optimal part of the road network. Within cutblocks, Hardy's method iterates over each cell in the harvest area and builds a road if it is not within skidding distance of an existing road, while we create landing points to serve as target locations for road construction.

<!--
TO DO: should this paragraph about speed and memory go in the results or discussion rather than methods?
-->
The MST method is expected to be closer to an optimal solution than iterative least cost path methods, and also more computationally costly. The Hardy QGIS plugin is much faster than our R package, and better able to handle large landscapes. The `igraph` package that we used to implement the Dijkstra algorithm requires the values of the cost raster to be read into memory in R which can lead to high memory usage on larger landscapes. See Appendix X for a more detailed exploration of the speed and memory requirements of the different methods.

@@

```{r proj-meth-fig, fig.cap="Comparison of minimum spanning tree (MST), dynamic least cost paths (DLCP), and the Hardy QGIS plugin road network projection algorithms. Landing locations were selected as the centroid of the cutblock, or a random or regularly spaced sample of points inside each cutblock (Table 2). Target cutblocks are outlined in grey, existing roads are black, projected roads are yellow, observed roads are red. Cost (1 ha resolution) is indicated by grey shading. The large four sided cutblock in the top right was added to more clearly show algorithm behaviour within cutblocks."}
knitr::include_graphics(here("analysis/figures/projection_methods_figure.png"))
```

@@

```{r aspat-perf-fig, fig.height=4, fig.width=fig_widths["two"], fig.cap="Proportional difference of projected mean from observed mean road metrics using different projection method variants within cutblocks, outside cutblocks, and overall (1 ha cost raster). Values greater than zero indicate overprojection and negative values indicate underprojection. Cutblocks not accessible via the observed road network were excluded from analysis. Performance varies among metrics (columns), projection method variants (colours), and within and outside cutblocks."}
# organize data for ggplot
doAspatPlot <- function(tab){
  meanTable_long <- tab %>% dplyr::select(-runTime, -order) %>% 
  pivot_longer(-c(sampleType, sampleDens, areaMean, method),
               names_to = "metric", values_to = "response")%>%
  mutate(sampleType = ifelse(sampleType == sampleDens, NA_character_, 
                             sampleType),
         sampleDens = paste(method, sampleType, sampleDens) %>% 
           factor(levels = c("mst NA centroid",
                             "NA NA observed",
                             "mst random low sample density",
                             "mst regular low sample density",
                             "mst random high sample density",
                             "mst regular high sample density",
                             "dlcp regular high sample density",
                             "NA NA klementQGIS",
                             "NA NA NA"),
                  labels = c("MST centroid", 
                             "Observed",
                             "MST random low density",
                             "MST regular low density",
                             "MST random high density",
                             "MST regular high density",
                             "DLCP regular high density",
                             "Hardy QGIS",
                             "Cutblocks only")),
         metric = factor(metric,
                         levels = c("roadDensityMean", "roadPresenceMean",
                                    "distanceToRoadMean",
                                    "forestryDisturbanceMean",
                                    "roadDisturbanceMean"),
                         labels = c("Road\ndensity", "Road\npresence",
                                    "Distance\nto road",
                                    "Forestry\ndisturbance\nfootprint",
                                    "Road\ndisturbance\nfootprint")))

@@

```{r proj-map-fig, fig.height=9, fig.width=fig_widths["two"], fig.cap="Comparison of the minimum spanning tree projection with regular high density sampling to the observed road network. The main map shows observed distance to the nearest road minus the projected distance. Insets highlight areas of interest with satellite imagery (Google 2022) for context: (a) Projected and observed road networks follow valleys but projected roads cross valley bottoms more frequently; (b) some open areas are misidentified as cutblocks; (c) projected methods do no mimic observed switchbacks."}

@@

```{r spat-perf-fig, fig.height=4.5, fig.width=fig_widths["two"], fig.cap="Aspatial variation of in performance among projection method variants (1 ha cost raster). F-measure is the harmonic mean of precision and sensitivity, precision is the proportion of predicted pixels that were observed and sensitivity is the proportion of observed pixels that were correctly predicted. Values closer to one indicate better performance. Cutblocks not accessible via the observed road network were excluded from analysis."}
doSpatPerf <- function(tab){
  matchData_sum <- tab %>% ungroup() %>%
    mutate(agreement = factor(agreement,
                              levels = c("False negative", "False positive",
                                         "Agree roaded","Agree roadless",
                                         "Pre-existing roads")),
           sampleDens = paste(method, sampleType, sampleDens) %>%
             str_replace("1e-06", "low density") %>%
             str_replace("1e-05", "high density") %>%
             str_replace("centroid low density", "centroid") %>%
             str_replace("klementQGIS NA", "klementQGIS") %>%
             factor(levels = c("mst centroid",
                               "mst random low density",
                               "mst regular low density",
                               "mst random high density",
                               "mst regular high density",
                               "dlcp regular high density",
                               "NA klementQGIS", 
                               "NA cutOnly NA"),
                    labels = c("MST centroid", 
                               "MST random low density",
                               "MST regular low density",
                               "MST random high density",
                               "MST regular high density",
                               "DLCP regular high density",
                               "Hardy QGIS", 
                               "Cutblocks only")),
           metric = factor(metric,
                           levels = c("roadPresence",
                                      "forestryDisturbance",
                                      "roadDisturbance"),
                           labels = c("Road presence",
                                      "Forestry disturbance\nfootprint",
                                      "Road disturbance\nfootprint"))) %>%
    group_by(metric, sampleDens, agreement) %>%
    summarise(count = sum(count), .groups = "drop_last") %>%
    mutate(perc = count/sum(count)*100)
